### 2015-03-11 src/4_OnlineLearning.tex
AVECM

Moreover, in case we receive data streams, ANN or SVM cannot introduce new information into the model without a re-training process, so we will have to use the same non-updated model until we decide to compute another one if it is possible.  Online learning algorithms allow one example at a time to be introduced into an existing model incrementally~\cite{vovk2005}. This is extremely important when the problem has large data streams and real-time forecasting must be done.  This is the most common scenario when we want to forecast a wide range of data such as stock prices and volatilities, electricity power, intrusion detection, web-mining, server load, etc.  Besides, many problems of high interest in machine learning can be treated as online ones and they can also use these types of algorithms.

The online learning framework was first introduced in the perceptron algorithm~\cite{rosenblatt58}. There are several other widely used online methods such as passive-aggressive~\cite{crammerETall2006}, stochastic gradient descent~\cite{zhang2004}, aggregating algorithm~\cite{vovk2001} and the second order perceptron~\cite{cesa-bianchi2005}.  In~\cite{cesa-bianchi2006} an in-depth analysis of online learning is provided.

>>>>

Moreover, in case we receive data streams, ANN or SVM cannot introduce new information into the model without a re-training process, so we will have to use the same non-updated model until we decide to compute another one if it is possible.  Online learning algorithms allow one example at a time to be introduced into an existing model incrementally \cite{vovk2005}. This is extremely important when the problem has large data streams and real-time forecasting must be done.  This is the most common scenario when we want to forecast a wide range of data such as stock prices and volatilities, electricity power, intrusion detection, web-mining, server load, etc.  Besides, many problems of high interest in machine learning can be treated as online ones and they can also use these types of algorithms.

The online learning framework was first introduced in the perceptron algorithm \cite{rosenblatt58}. There are several other widely used online methods such as passive-aggressive \cite{crammerETall2006}, stochastic gradient descent \cite{zhang2004}, aggregating algorithm \cite{vovk2001} and the second order perceptron \cite{cesa-bianchi2005}.  In \cite{cesa-bianchi2006} an in-depth analysis of online learning is provided.

### 2015-03-11 src/5_1_Proposal.tex
AVECM

\chapter{Online VECM proposal}
Financial time series are known for their non-stationary behaviour. However,
sometimes they exhibit some stationary linear combinations. When this happens,
it is said that those time series are cointegrated.The Vector Error Correction
Model (VECM) is an econometric model which characterizes the joint dynamic
behaviour of a set of cointegrated variables in terms of forces pulling towards
equilibrium. In this study, we propose an Online VEC model (OVECM) which
optimizes how model parameters are obtained using a sliding window of the most
recent data. Our proposal also takes advantage of the long-run relationship
between the time series in order to obtain improved execution times. Our
proposed method is tested using four foreign exchange rates with a frequency of
1-minute, all related to the USD currency base. OVECM is compared with VECM and
ARIMA models in terms of forecasting accuracy and execution times. We show that
OVECM outperforms ARIMA forecasting and enables execution time to
be reduced considerably while maintaining good accuracy levels compared with
VECM.

@@

Some financial models, such as  vector error correction (VEC), take advantage of
this property and describe the joint behaviour of several cointegrated
financial instruments.

VEC is a special case of the vector autorregressive model (VAR). VAR is a linear
model which express future values in terms of its own history and other
variables past values. If cointegration exists, VEC model allows to introduce an
error correction term due to its cointegration estimation error.

VEC as well as VAR model parameters are obtained using ordinary least squares
method (OLS). Since OLS involves many calculations, the parameter estimation
method is computationally expensive when the number of past values and
observations considered increases. Moreover, OLS is an ill-posed problem which
admits an infinite number of solutions.

@@

There are several popular online methods such as
perceptron~\cite{rosenblatt58}, passive-aggressive~\cite{crammerETall2006},
stochastic gradient descent~\cite{zhang2004}, aggregating
algorithm~\cite{vovk2001} and the second order
perceptron~\cite{cesa-bianchi2005}.  In~\cite{cesa-bianchi2006}, an in-deph
analysis of online learning is provided.  In particular, the {\em aggregating
algorithm for regression} (AAR) method is a recursive formulation of RR
suitable for using in an online context.

@@

\subsection{Online VECM} \label{sec:proposal}

@@

OVECM also implements matrix optimizations in order to reduce execution time,
such as updating matrices with new data, removing old data and introducing new
cointegration vectors.

@@

\begin{itemize}
\item The function \texttt{getJohansen} returns cointegration vectors given by
the Johansen method considering the trace statistic test at 95\%
significance level.
\item The function \texttt{vecMatrix} returns the matrices~(\ref{eq:vecA})
and~(\ref{eq:vecB}) that allows VECM to be solved.
\item The function \texttt{vecMatrixOnline} returns the
matrices~(\ref{eq:vecA}) and~(\ref{eq:vecB}) aggregating new data and removing
the old one, avoiding calculation of the matrix $\mathbf{A}$ from scratch.
\item Out-of-sample outputs are saved in the variables 
$\mathbf{Y}_{\text{true}}$ and $\mathbf{Y}_{\text{pred}}$.
\item The model is solved using OLS.
\item In-sample outputs are saved in the variables $\Delta
\mathbf{y}_{\text{true}}$ and $\Delta \mathbf{y}_{\text{pred}}$.
\item The function \texttt{mape} gets the in-sample MAPE for the $l$ time
series.
\item Cointegration vectors are obtained at the beginning and when they are required to be updated. This updating is decided based on the in-sample MAPE of the last $n$ inputs. The average of all
MAPEs are stored in the variable $e$. If the average of MAPEs
($\texttt{mean}(e)$) is above a certain error given by the mean\_error threshold, cointegration vectors are updated.
\item If new cointegration vectors are required, the function
\texttt{vecMatrixUpdate} only updates the corresponding columns of matrix
$\mathbf{A}$ affected by those vectors (see equation~\ref{eq:vecA}).
\end{itemize}

\begin{algorithm}[ht]
\begin{algorithmic}[1]
\REQUIRE $\,$ \\
$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
$p$: number of past values \\
$L$: sliding window size ($L<N$) \\
$\text{mean\_error}$: MAPE threshold \\
$n$: interpolation points to obtain MAPE \\
\ENSURE  $\,$ \\
$\{\mathbf{y}_{\text{pred}}[L+1],\dots,\mathbf{y}_{\text{pred}}[N]\}$: model predictions 
\FOR { $i =0$ to $N-L$ }
    \STATE $\mathbf{y}_i \gets \mathbf{y}[i:i+L]$
	\IF {i = 0 }
	    \STATE{$v \gets \texttt{getJohansen}(\mathbf{y}_i,p)$}
	    \STATE{$[\mathbf{A} \quad \mathbf{B}] \gets
        \texttt{vecMatrix}(\mathbf{y}_i,p,v)$}
	\ELSE
	    \STATE{$[\mathbf{A} \quad \mathbf{B}] \gets
        \texttt{vecMatrixOnline}(\mathbf{y}_i,p,v,\mathbf{A},\mathbf{B})$}
        \STATE $\Delta \mathbf{Y}_{\text{pred}}[i] \gets \mathbf{A}[-1,:] \times \mathbf{X}$
    \ENDIF
    \STATE $\mathbf{X} \gets \texttt{OLS} (\mathbf{A},\mathbf{B})$
    \STATE $e \gets \texttt{mape}(\mathbf{B}[-n,:],\mathbf{A}[-n,:] \times \mathbf{X})$
    \IF {$\texttt{mean}(e) > \text{mean\_error}$}
	    \STATE{$v \gets \texttt{getJohansen}(\mathbf{y}_i,p)$}
	    \STATE{$\mathbf{A} \gets
        \texttt{vecMatrixUpdate}(\mathbf{y}_i,p,v,\mathbf{A})$}
        \STATE $\mathbf{X} \gets \texttt{OLS} (\mathbf{A},\mathbf{B})$
    \ENDIF
\ENDFOR
\STATE $\mathbf{Y}_{\text{true}} \gets \mathbf{Y}[L+1:N]$
\STATE $\mathbf{Y}_{\text{pred}}\gets 
\mathbf{Y}[L:N-1] +\Delta \mathbf{Y}_{\text{pred}}$
\end{algorithmic}
\caption{OVECM: Online VECM}
\label{alg:proposal}
\end{algorithm}

@@

    \begin{tikzpicture}[node distance=2cm]
    %\node (start) [startstop] {Start};
    %\node (in1) [io, below of=start] {Input};
    \node (in1) [io] {Input: y, p, L, $\epsilon$, n};
    \node (pro1) [process, below of=in1] {$y_i = y[i:i+L]$};
    \node (dec1) [decision, below of=pro1, yshift=-0.5cm] {i = 0};
    \node (pro2a) [process, below of=dec1, yshift=-0.5cm] {$[\mathbf{A} \quad \mathbf{B}] =\texttt{ovecm}(\mathbf{y}_i,p,v,\mathbf{A},\mathbf{B})$
                                                        $\Delta \mathbf{Y}_{\text{true}}[i] = \mathbf{B}[-1,:]$
                                                        $\Delta \mathbf{Y}_{\text{pred}}[i] = \mathbf{A}[-1,:] \times \mathbf{X}$};

    \node (pro2b) [process, right of=dec1, xshift=4cm] {$v = \texttt{Johansen}(\mathbf{y}_i,p)$ $[\mathbf{A} \quad 
                                                                \mathbf{B}] = \texttt{vecm}(\mathbf{y}_i,p,v)$};
    \node (pro3) [process, below of=pro2a, yshift=-0.5cm] {$\mathbf{X} = \text{OLS} (\mathbf{A},\mathbf{B})$ \\
                                                        $\mathbf{y}_{\text{true}}[i] = \mathbf{y}_i[-n,:]$ \\
                                                        $\mathbf{y}_{\text{pred}}[i] = \mathbf{A}[-n,:] \times \mathbf{X} +$ \\$ \mathbf{y}_i[-n-1,:-1]$ \\
                                                        $e = \texttt{mape}(\mathbf{y}_{\text{true}}, \mathbf{y}_{\text{pred}})$};

    \node (dec2) [decision, below of=pro3, yshift=-0.5cm] { $e > \epsilon $ };

    \node (pro4a) [process, right of=dec2, xshift=4cm] {$v = \texttt{Johansen}(\mathbf{y}_i,p)$ \\
                                                        $\mathbf{A} = \texttt{uvecm}(\mathbf{y}_i,p,v,\mathbf{A})$ \\
                                                        $\mathbf{X} = \texttt{OLS} (\mathbf{A},\mathbf{B})$};

    \node (pro5) [process, left of=pro2a, xshift=-4cm] {$ i++ $};

    \node (dec3) [decision, below of=dec2, yshift=-0.5cm] { $i > it $ };

    \node (out1) [io, below of=dec3] {Output: $\Delta \mathbf{y}_{true}, \Delta \mathbf{y}_{pred}$};

    %\node (stop) [startstop, below of=out1] {Stop};

    %\draw [arrow] (start) -- (in1);
    \draw [arrow] (in1) -- (pro1);
    \draw [arrow] (pro1) -- (dec1);
    \draw [arrow] (dec1) -- node[anchor=east] {no} (pro2a);
    \draw [arrow] (dec1) -- node[anchor=south] {si} (pro2b);
    \draw [arrow] (pro2a) -- (pro3);
    \draw [arrow] (pro2b) |- (pro3);
    \draw [arrow] (pro3) -- (dec2);
    \draw [arrow] (dec2) --  node[anchor=south] {si} (pro4a);
    \draw [arrow] (pro4a) |- (dec3);

    \draw [arrow] (dec2) --  node[anchor=east] {no} (dec3);
    \draw [arrow] (dec3.west) -| node[anchor=east] {no} (pro5);
    \draw [arrow] (pro5) |- (pro1);
    \draw [arrow] (dec3) --  node[anchor=east] {si} (out1);
    %\draw [arrow] (out1) -- (stop);
    \end{tikzpicture}

>>>>

\chapter{Online VECM proposal}
Financial time series are known for their non-stationary behaviour. However,
sometimes they exhibit some stationary linear combinations. When this happens,
it is said that those time series are cointegrated. The Vector Error Correction
Model (VECM) is an econometric model which characterises the joint dynamic
behaviour of a set of cointegrated variables in terms of forces pulling towards
equilibrium. In this study, we propose an Online VEC model (OVECM) which
optimises how model parameters are obtained using a sliding window of the most
recent data. Our proposal also takes advantage of the long-run relationship
between the time series in order to obtain improved execution times. Our
proposed method is tested using four foreign exchange rates with a frequency of
1-minute, all related to the USD currency base. OVECM is compared with VECM and
ARIMA models in terms of forecasting accuracy and execution times. We show that
OVECM outperforms ARIMA forecasting and enables execution time to
be reduced considerably while maintaining good accuracy levels compared with
VECM.

@@

Some financial models, such as  vector error correction model (VECM), take advantage of
this property and describe the joint behaviour of several cointegrated
financial instruments.

VECM is a special case of the vector autoregressive model (VAR). VAR is a linear
model which express future values in terms of its own history and other
variables past values. If cointegration exists, VECM allows to introduce an
error correction term due to its cointegration estimation error.

VECM as well as VAR model parameters are obtained using ordinary least squares
method (OLS). Since OLS involves many calculations, the parameter estimation
method is computationally expensive when the number of past values and
observations considered increases. Moreover, OLS is an ill-posed problem which
admits an infinite number of solutions.

@@

@@

@@

OVECM also implements matrix optimisations in order to reduce execution time,
such as updating matrices with new data, removing old data and introducing new
cointegration vectors.

@@

\begin{itemize}
\item The function \texttt{getJohansen} returns cointegration vectors given by
the Johansen method considering the trace statistic test at 95\%
significance level.
\item The function \texttt{vecMatrix} returns the matrices~(\ref{eq:vecA})
and~(\ref{eq:vecY}) that allows VECM to be solved.
\item The function \texttt{vecMatrixOnline} returns the
matrices~(\ref{eq:vecA}) and~(\ref{eq:vecY}) aggregating new data and removing
the old one, avoiding calculation of the matrix $\mathbf{A}$ from scratch.
\item Out-of-sample outputs are saved in the variables 
$\mathbf{Y}_{\text{true}}$ and $\mathbf{Y}_{\text{pred}}$.
\item The model is solved using OLS.
\item In-sample outputs are saved in the variables $\Delta
\mathbf{y}_{\text{true}}$ and $\Delta \mathbf{y}_{\text{pred}}$.
\item The function \texttt{mape} gets the in-sample MAPE for the $l$ time
series.
\item Cointegration vectors are obtained at the beginning and when they are required to be updated. This updating is decided based on the in-sample MAPE of the last $n$ inputs. The average of all
MAPEs are stored in the variable $e$. If the average of MAPEs
($\texttt{mean}(e)$) is above a certain error given by the mean\_error threshold, cointegration vectors are updated.
\item If new cointegration vectors are required, the function
\texttt{vecMatrixUpdate} only updates the corresponding columns of matrix
$\mathbf{A}$ affected by those vectors (see equation~\ref{eq:vecA}).
\end{itemize}

\begin{algorithm}[ht]
\begin{algorithmic}[1]
\REQUIRE $\,$ \\
$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
$p$: number of past values \\
$L$: sliding window size ($L<N$) \\
$\text{mean\_error}$: MAPE threshold \\
$n$: interpolation points to obtain MAPE \\
\ENSURE  $\,$ \\
$\{\mathbf{y}_{\text{pred}}[L+1],\dots,\mathbf{y}_{\text{pred}}[N]\}$: model predictions 
\FOR { $i =0$ to $N-L$ }
    \STATE $\mathbf{y}_i \gets \mathbf{y}[i:i+L]$
	\IF {i = 0 }
	    \STATE{$v \gets \texttt{getJohansen}(\mathbf{y}_i,p)$}
	    \STATE{$[\mathbf{A} \quad \mathbf{Y}] \gets
        \texttt{vecMatrix}(\mathbf{y}_i,p,v)$}
	\ELSE
	    \STATE{$[\mathbf{A} \quad \mathbf{Y}] \gets
        \texttt{vecMatrixOnline}(\mathbf{y}_i,p,v,\mathbf{A},\mathbf{Y})$}
        \STATE $\Delta \mathbf{Y}_{\text{pred}}[i] \gets \mathbf{A}[-1,:] \times \mathbf{X}$
    \ENDIF
    \STATE $\mathbf{X} \gets \texttt{OLS} (\mathbf{A},\mathbf{Y})$
    \STATE $e \gets \texttt{mape}(\mathbf{Y}[-n,:],\mathbf{A}[-n,:] \times \mathbf{X})$
    \IF {$\texttt{mean}(e) > \text{mean\_error}$}
	    \STATE{$v \gets \texttt{getJohansen}(\mathbf{y}_i,p)$}
	    \STATE{$\mathbf{A} \gets
        \texttt{vecMatrixUpdate}(\mathbf{y}_i,p,v,\mathbf{A})$}
        \STATE $\mathbf{X} \gets \texttt{OLS} (\mathbf{A},\mathbf{Y})$
    \ENDIF
\ENDFOR
\STATE $\mathbf{Y}_{\text{true}} \gets \mathbf{Y}[L+1:N]$
\STATE $\mathbf{Y}_{\text{pred}}\gets 
\mathbf{Y}[L:N-1] +\Delta \mathbf{Y}_{\text{pred}}$
\end{algorithmic}
\caption{OVECM: Online VECM}
\label{alg:proposal}
\end{algorithm}

A pseudocode of the algorithm \ref{alg:proposal} is detailed in the appendix \ref{fig:pseudocodigo}.

@@
